---
title: "coop-careers-data-proj-1"
author: "allen phu"
date: "2025-09-16"
output:
  html_document: default
  pdf_document: default
---

# Initial Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

```{r}
# load libraries
library(tidyverse)
library(lubridate)
library(tidymodels)
library(writexl)
library(showtext)

# load the 'Lato' font from Google Fonts and give it the nickname 'lato'
font_add_google("Lato", "lato")

# tell R to use the showtext package for all plots
showtext_auto()
```

# Data Load

```{r}
# load raw data
data <- read.csv("raw_data/nyc_restaurant_inspection.csv")
```

```{r}
# peek at columns, what their values are
glimpse(data)
```

# Data Cleaning

Ok, so what does **drop_na()** do in R? From direct documentation: drops rows where any column (in this case, the entire dataset/dataframe) contains a missing value. So, in this case, anything with NA.

## wtf is NA?

NA is different from "empty"; **NA** is considered "separate" and "missing", **"empty"** or **""** is a separate thing completely.

So first, we need to add row numbers; this is pre-processing for later.

```{r create-id-column}
# throwing a row number on the original data so we can track stuff later
data_with_id <- data |>
  mutate(original_row = row_number())
```

My prof from undergrad described pivot_longer as this: *"For when some of the column names are not names of variables, but values of a variable"*

```{r find-na-locations}
# we want to find where "NA" values are
na_locations <- data_with_id |>
  # first, just grab the rows that have an NA somewhere in them
  filter(if_any(-original_row, is.na)) |>
  # make everything a character so the pivot doesn't give us errors
  mutate(across(-original_row, as.character)) |>
  # convert our table to a long/increase rows and 
  pivot_longer(
    cols = -original_row,
    names_to = "column_name",
    values_to = "value"
  ) |>
  # now, just keep the lines that are NA
  filter(is.na(value))

# look at list
# print(na_locations)
```

# Correct Data Types

However, we also have empty columns, so we can find that too.

`{# {r find-empty-locations} # # ok, same thing but for the empty "" values # empty_string_locations <- data_with_id |> #   # grab any row that has an empty "" in a character column #   filter(if_any(where(is.character),function(x) x == "")) |> #   # again, making everything a character to avoid that pivot error #   mutate(across(-original_row, as.character)) |> #   # convert table into "longer" table #   pivot_longer( #     cols = -original_row, #     names_to = "column_name", #     values_to = "value" #   ) |> #   # just find the rows that are actually empty strings #   filter(value == "") #  # # print out list of empty value locations # print(empty_string_locations)`

```{r}
# actually clean the data here, with the correct order of operations
cleaned_data <- data |>
  # remove rows with duplicate CAMIS/Date/Code
  distinct(CAMIS, INSPECTION.DATE, VIOLATION.CODE, .keep_all = TRUE) |> 
  # make the date columns "actual dates" first
  mutate(
    INSPECTION.DATE = mdy(INSPECTION.DATE),
    GRADE.DATE = mdy(GRADE.DATE),
    RECORD.DATE = mdy(RECORD.DATE)) |>
  # now, get rid of rows with NA and empty strings
  # why? mdy() will automatically convert "" into an NA, then drop_NA() will remove it/drop the row
  drop_na() |>
  filter(!if_any(where(is.character), function(x) x == ""))
```

```{r}
# use count() to find which combinations have a count (n) greater than 1
duplicate_summary <- cleaned_data |>
  count(CAMIS, INSPECTION.DATE, sort = TRUE) |>
  filter(n > 1)

# print the summary of which records are duplicated
# print(duplicate_summary)
```

```{r glimpse-and-summary-cleaned-data}
# quick check to see if our cleaned data looks right
glimpse(cleaned_data)
# summary(cleaned_data)
```

```{r}
# Step 1: Create the data table for the plot
top_10_critical <- cleaned_data |> 
  filter(CRITICAL.FLAG == "Critical") |> 
  count(VIOLATION.CODE, sort = TRUE) |> 
  slice_head(n = 10)

# Now, print the table so you can see the result of slice_head()
print(top_10_critical)
```

```{r barchart-high-def, fig.width=10, fig.height=7, dpi=300}
top_10_critical |>
  mutate(VIOLATION.CODE = fct_reorder(VIOLATION.CODE, n)) |>
  ggplot(aes(x = n, y = VIOLATION.CODE)) +
  geom_col(fill = "firebrick") +
  geom_text(aes(label = n), hjust = -0.2, size = 16) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "Top 10 Critical Violation Codes",
    x = "Total Count of Violations",
    y = "Violation Code"
  ) +
  
  # FIX 1: Increase the base font size for the whole plot
  theme_minimal(base_size = 60) +
  
  # FIX 2: Remove ALL grid lines (both vertical and horizontal)
  theme(
    text = element_text(family = "lato"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
```

# Write to CSV, XLSX

`{# {r} # # save data # write_csv(cleaned_data, "cleaned_nyc_inspections.csv") # write_xlsx(cleaned_data, "cleaned_nyc_inspections.xlsx")`
